---
title: "Get started with datadiffr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with datadiffr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datadiffr)
```

datadiffr compares two data frames and reports the differences the way a code
review shows a diff: cell by cell, with a few unchanged rows around each change
for context. This vignette walks through the core workflow — comparing frames,
controlling context, matching rows by key, tolerating small numeric
differences, and rendering an HTML report.

## A first comparison

Start with two versions of the same small table. `compare_data()` returns a
`datadiff_result` — an object describing what it found. It prints a one-line
summary above a table of the differences, which is handy when you are working in
a plain console rather than the HTML report.

```{r}
before <- data.frame(
  id    = 1:6,
  name  = c("Ana", "Ben", "Cai", "Dana", "Eve", "Finn"),
  score = c(88, 91, 75, 82, 95, 60)
)
after <- data.frame(
  id    = 1:6,
  name  = c("Ana", "Ben", "Cai", "Dana", "Eve", "Finn"),
  score = c(88, 91, 79, 82, 95, 72)
)

compare_data(before, after)
```

The object's `$kind` field names what kind of difference was found:

```{r}
diff <- compare_data(before, after)
diff$kind
```

`"value"` means the two frames share the same columns but some values differ.
(The other kinds are `"identical"`, when the frames match, and `"schema"`, when
the columns themselves differ — see [When the columns themselves
differ](#when-the-columns-themselves-differ) below.) For a value diff, the
differing rows are in `$rows`, a tidy data frame carrying a few bookkeeping
columns alongside your data:

- `.row` — the row number the difference came from.
- `.join_type` — whether the row is in `"x"`, `"y"`, or `"both"`.
- `.diff_type` — whether the row is a `"diff"` or a `"context"` row.
- `.source` — for diff rows, whether it is the `"x"` (before) or `"y"` (after)
  version; `NA` for context rows.

Each changed row appears twice — the before value above the after value — so you
can read the change top to bottom, like a unified diff.

## Controlling context

By default datadiffr surrounds each change with three unchanged rows before and
after. `context_rows` takes a length-two vector — rows before, rows after — so
you can widen, narrow, or drop the context entirely.

```{r}
# No context: only the rows that actually changed
compare_data(before, after, context_rows = c(0L, 0L))
```

```{r}
# One row of context on each side
compare_data(before, after, context_rows = c(1L, 1L))
```

You can also restrict which columns are shown as context with `context_cols`,
using tidyselect syntax (for example `context_cols = c(id, name)`).

## Matching rows by key

The examples above match rows by position. When rows can move — a sort changed,
rows were inserted or deleted — match on one or more key columns instead with
`by =`. Keys must exist in both frames and identify rows uniquely; the output is
ordered by the key and always includes it.

```{r}
# `after` is shuffled and has a new row (id 7) and a dropped row (id 3)
after_shuffled <- data.frame(
  id    = c(2L, 1L, 4L, 5L, 6L, 7L),
  name  = c("Ben", "Ana", "Dana", "Eve", "Finn", "Gwen"),
  score = c(91, 88, 82, 95, 72, 70)
)

compare_data(before, after_shuffled, by = "id", context_rows = c(0L, 0L))
```

Rows present in only one frame are always reported: id 3 shows up as removed
(in `x` only) and id 7 as added (in `y` only).

## Tolerating small numeric differences

Floating-point data is rarely bit-identical. `tolerance` treats numeric values
within a threshold as equal, so tiny differences don't flood the report.

```{r}
measured  <- data.frame(id = 1:3, value = c(1.0000, 2.0000, 3.0000))
recomputed <- data.frame(id = 1:3, value = c(1.0001, 2.5000, 3.0000))

# Exact-ish comparison flags both id 1 and id 2
compare_data(measured, recomputed, context_rows = c(0L, 0L))

# A tolerance of 0.01 keeps only the genuine change (id 2)
compare_data(measured, recomputed, tolerance = 0.01, context_rows = c(0L, 0L))
```

The same `tolerance` argument is available on `diffdata()` and `is_equal()`, and
it applies on the natural scale for dates (days) and datetimes (seconds).

## When the columns themselves differ

datadiffr compares values, so it expects the two frames to share the same
columns with compatible types. If they don't, `compare_data()` doesn't error —
it returns a `"schema"` result reporting the *column* differences instead of
diffing the values.

```{r}
x <- data.frame(id = 1:3, value = 1:3, note = c("a", "b", "c"))
y <- data.frame(id = 1:3, value = c(1.5, 2.5, 3.5))

schema_diff <- compare_data(x, y)
schema_diff$kind
```

For a schema result the column differences are in `$columns` — the same table
`compare_columns()` returns directly:

```{r}
schema_diff$columns
```

Here `value` changed type (integer to double) and `note` exists only in `x`.
Because the kind is always one of `"value"`, `"identical"`, or `"schema"`,
calling code can branch on `$kind` and read `$rows` or `$columns` accordingly.

## Rendering an HTML report

`diffdata()` runs the comparison and opens the result as a styled, self-contained
HTML report — changed rows colored red (before) and green (after), with context
rows grayed out. In an interactive session it opens in the RStudio viewer or your
browser:

```{r eval = FALSE}
diffdata(before, after)
```

To save the report instead of opening it, pass `output_file`:

```{r eval = FALSE}
diffdata(before, after, output_file = "diff-report.html")
```

`diffdata()` defaults to `max_differences = 10` to keep reports fast to render;
raise it if you need every difference in the report. `compare_data()`, which is
meant for programmatic use, reports everything by default.

If you already have a diff from `compare_data()`, render it directly with
`render_diff()`:

```{r eval = FALSE}
diff <- compare_data(before, after)
render_diff(diff)
```

## Coming from dataCompareR?

datadiffr also ships a clean-room drop-in replacement for the archived
dataCompareR package — `rCompare()` and friends. See
`vignette("migrating-from-datacomparer")`.
