---
title: "Migrating from dataCompareR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from dataCompareR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

dataCompareR was archived by CRAN in February 2026 and its repository is
read-only. datadiffr covers the same workflow — compare two data frames,
summarize the result, extract the mismatches, save a report — with a
smaller, snake_case API. This guide maps each dataCompareR call to its
datadiffr equivalent.

## Call-by-call mapping

| dataCompareR | datadiffr |
|---------------------------------------------|--------------------------------------------------|
| `rCompare(dfA, dfB)` | `compare_data(x, y)` |
| `rCompare(dfA, dfB, keys = "id")` | `compare_data(x, y, by = "id")` |
| `rCompare(dfA, dfB, roundDigits = 2)` | `compare_data(x, y, tolerance = 0.005)` |
| `rCompare(dfA, dfB, trimChars = TRUE)` | trim first: `mutate(across(where(is.character), trimws))` |
| `summary(comparison)` | `summary(result)` |
| `comparison$mismatches` | `get_cell_differences(result)` |
| `generateMismatchData(comparison, dfA, dfB)` | `get_differences(result)` |
| `saveReport(comparison, "report")` | `render_diff(result, output_file = "report.html")` |

A worked example:

```{r}
library(datadiffr)

x <- data.frame(id = 1:4, score = c(10, 20, 30, 40))
y <- data.frame(id = 1:4, score = c(10, 25, 30, 45))

result <- compare_data(x, y, by = "id")
summary(result)

get_cell_differences(result)
```

## Differences worth knowing

* **Column mismatches.** dataCompareR silently compared the intersection of
  columns, matching names case-insensitively. datadiffr treats column
  differences as a result in their own right: `compare_data()` returns a
  `"schema"` result describing them (see `compare_columns()`) and does not
  compare values. To compare anyway, select the shared columns first.
* **Extra rows under positional matching.** dataCompareR truncated the
  longer frame; datadiffr reports the extra rows as differences.
* **Rounding vs tolerance.** `roundDigits = k` is approximately
  `tolerance = 0.5 * 10^-k`. Rounding can still call two nearly-equal
  values different when they round apart; tolerance compares distance
  directly.
* **Reports.** `render_diff()` and `diffdata()` produce a styled HTML diff
  with context rows rather than a text report; `summary(result)` covers the
  console.
