---
title: "Migrating from dataCompareR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from dataCompareR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

dataCompareR was archived by CRAN in February 2026 and its repository is
read-only. If `rCompare()` is part of your workflow, datadiffr provides a
drop-in replacement for the package's entire exported surface —
`rCompare()`, `print()`, `summary()`, `saveReport()`, and
`generateMismatchData()` — reimplemented from scratch (no dataCompareR
code is reused) and verified field-by-field against dataCompareR itself
in datadiffr's test suite.

In most scripts, migration is one line:

```{r, eval = FALSE}
# library(dataCompareR)
library(datadiffr)
```

## Quick start

```{r}
library(datadiffr)

orders_old <- data.frame(
  order_id = 1:5,
  amount = c(10, 20.5, 30, 45, 50),
  status = c("open", "open", "closed", "open", "closed")
)
orders_new <- data.frame(
  order_id = c(1:4, 6L),
  amount = c(10, 21.0, 30, 45, 60),
  status = c("open", "paid", "closed", "open", "open")
)

cmp <- rCompare(orders_old, orders_new, keys = "order_id")
cmp
```

The full report, as before:

```{r}
summary(cmp)
```

Everything downstream works the same way:

```{r}
mismatches <- generateMismatchData(cmp, orders_old, orders_new)
mismatches$orders_old_mm
```

```{r, eval = FALSE}
saveReport(cmp, reportName = "orders", reportLocation = ".")
```

## The compatibility surface

| dataCompareR                  | datadiffr             | Notes                                        |
|-------------------------------|-----------------------|----------------------------------------------|
| `rCompare(dfA, dfB, keys, roundDigits, mismatches, trimChars)` | Same | Plus a `tolerance` extension (see below)     |
| `print(x, nVars, nObs, verbose)` | Same               | Same console format                          |
| `summary(x, mismatchCount)`   | Same                  | Same fields; deterministic detail tables     |
| `saveReport(...)`             | Same                  | Same arguments, rendered with rmarkdown      |
| `generateMismatchData(x, dfA, dfB)` | Same            | Same validation and return shape             |

The returned object has class
`c("datadiff_compare", "dataCompareRobject")` and the same six fields
(`meta`, `colMatching`, `rowMatching`, `cleaninginfo`, `mismatches`,
`matches`) with the same shapes, so code that inspects the object —
`cmp$mismatches$AMOUNT`, `cmp$rowMatching$inboth`, and so on — keeps
working.

The comparison semantics are preserved, including the corners:

* Column names are cleaned with `make.names()` and matched
  case-insensitively; shared columns are reported upper-case and sorted.
* Without `keys`, rows are compared by position and the longer frame is
  truncated (recorded in `rowMatching$inA$indices_removed`).
* With `keys`, key columns must exist in both frames (checked
  case-sensitively against the cleaned names) and must identify rows
  uniquely in each frame.
* `NA` matches `NA` and `NaN` matches `NaN`, but `NA` does not match
  `NaN`.
* Factors are compared as character, and the coercion is recorded in
  `cleaninginfo`.
* Columns whose classes differ between the frames (for example integer
  vs double) are reported as mismatching on every row, with types shown
  and a blank difference.
* The `mismatches` cap **errors** when exceeded — it does not truncate —
  exactly as dataCompareR behaved.

## Known differences

These are deliberate and documented; none change the object contract:

* **Detail tables are deterministic.** Where dataCompareR *sampled* rows
  for `summary()` detail tables beyond `mismatchCount`, datadiffr sorts
  by decreasing absolute difference and keeps the first rows. Reruns
  produce identical reports.
* **Metadata references datadiffr.** `summary()$version` is the
  datadiffr version, and report headers say "With datadiffr version …".
* **Clear validation errors.** Inputs must be data frames (dataCompareR
  silently coerced some malformed inputs into empty comparisons), and
  error messages are worded differently, though they fire in the same
  situations.

## Extensions

`rCompare()` gains a `tolerance` argument, appended after the original
arguments so positional calls behave identically. dataCompareR's only
fuzziness mechanism was `roundDigits` (round both sides, then compare
exactly), which fails on values that round in different directions:

```{r}
a <- data.frame(x = 1.14)
b <- data.frame(x = 1.16)
# rounds to 1.1 vs 1.2: reported as a mismatch
length(rCompare(a, b, roundDigits = 1)$mismatches)
# absolute tolerance: equal
length(rCompare(a, b, tolerance = 0.05)$mismatches)
```

The default `tolerance = 0` keeps faithful exact comparison.

The second extension is `render_diff()`: datadiffr's styled HTML diff
report, straight from the comparison object.

```{r, eval = FALSE}
render_diff(cmp)                          # opens in the viewer
render_diff(cmp, output_file = "cmp.html") # or saves to a file
```

## Graduating to the native API

The compatibility layer is a stable endpoint — you can stay on it. But
datadiffr's native API adds things `rCompare()` cannot express:

```{r, eval = FALSE}
diffdata(
  orders_old,
  orders_new,
  by = "order_id",        # key-based matching, like rCompare keys
  context_rows = c(2, 2), # unified-diff style context around changes
  context_cols = c(order_id, amount),
  tolerance = 0.01
)
```

`diffdata()` compares, then opens the HTML diff report with added rows
in green, removed rows in red, and unchanged context rows for
orientation. `compare_data()` returns the underlying diff as a data
frame for programmatic use.

## Troubleshooting

* **"Key … was not found in both data frames"** — key names are checked
  against the `make.names()`-cleaned column names, case-sensitively:
  `keys = "ID"` does not find a column named `id`. This matches
  dataCompareR.
* **"Detected N mismatches, which exceeds the `mismatches` cap"** — the
  cap is a guard, not a limit; raise it or drop the argument, as with
  dataCompareR.
* **`generateMismatchData()` says a frame was not part of the original
  comparison** — like dataCompareR, frames are matched by the *names*
  they were passed under at `rCompare()` time; pass the same variables.
* **`render_diff()` asks for flexdashboard** — the HTML diff report
  needs the suggested flexdashboard package:
  `install.packages("flexdashboard")`.
